---
title: "Horizon Plot"
author: "Shannon Lee"
date: "February 24, 2024"
date-modified: "last-modified"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
output: html_document
css: styles.css
execute: 
  eval: true
  echo: true
  warning: false
---

# **20. Time on the Horizon: ggHoriPlot methods**

## **20.2 Getting started**

Before getting start, make sure that **ggHoriPlot** has been included in the `pacman::p_load(...)` statement above.

```{r}
pacman::p_load(ggHoriPlot, ggthemes, tidyverse)
```

### **20.2.1 Step 1: Data Import**

For the purpose of this hands-on exercise, [Average Retail Prices Of Selected Consumer Items](https://tablebuilder.singstat.gov.sg/table/TS/M212891) will be used.

Use the code chunk below to import the AVERP.csv file into R environment.

```{r}
averp <- read_csv("../../data/AVERP.csv") %>%
  mutate(`Date` = dmy(`Date`))
```

### **20.2.2 Step 2: Plotting the horizon graph**

Next, the code chunk below will be used to plot the horizon graph.

```{r}
averp %>% 
  filter(Date >= "2018-01-01") %>%
  ggplot() +
  geom_horizon(aes(x = Date, y=Values), 
               origin = "midpoint", 
               horizonscale = 6)+
  facet_grid(`Consumer Items`~.) +
    theme_few() +
  scale_fill_hcl(palette = 'RdBu') +
  theme(panel.spacing.y=unit(0, "lines"), strip.text.y = element_text(
    size = 5, angle = 0, hjust = 0),
    legend.position = 'none',
    axis.text.y = element_blank(),
    axis.text.x = element_text(size=7),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_blank()
    ) +
    scale_x_date(expand=c(0,0), date_breaks = "3 month", date_labels = "%b%y") +
  ggtitle('Average Retail Prices of Selected Consumer Items (Jan 2018 to Dec 2022)')
```